 

### 2021-05-03 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

The strength of this buffering varies with foliage characteristics mentioned above, and water availability [@davis_microclimatic_2019]. Diurnally, tropical, temperate broadleaf, mixed-conifer and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory respectively, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al. 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; Rambo and North, 2009]. In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower understory peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with LAI and canopy cover [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. On the other hand, open-forests with sparse stands, have lower buffering capacity, thus greater vertical air mixing, greater light permeability through canopy layers and sometimes greater soil heating [@ramboCanopyMicroclimateResponse2009]. Therefore, they maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009]. During the night, pine forests have shown to maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a]. Canopy roughness also affects buffering, turbulent air flow and the canopy boundary layer. As a result, taller trees up to 20m in an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature in the understory. However, beyond a certain tree height, temperature and VPD offset plateaus [@juckerCanopyStructureTopography2018; @hardwickRelationshipLeafArea2015].

@@

While exposed sun leaves are equipped to thermoregulate more efficiently than shaded leaves, they are largely constrained by stomatal conductance; and therefore their ability to photosynthesize as well. Diurnally, stomtal conductance increases with height and is greater in the upper canopy compared to shaded canopies during the day [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010]. During midday, stomtal closure occurs, when evaporative demand exceeds water transport capacity in exposed leaves [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004]. Thus, stomatal limitation is greater and steeper in exposed canopies than shade leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020]. Similarly, max transpiration rate increases with canopy height [@kosugiSeasonalFluctuationsTemperature2006]. Stomatal apeture, also controls heat dissapation through evaporative cooling [Taiz and Zeiger], where understory leaves tend to open their stomata wider than upper canopy leaves [@kosugiSeasonalFluctuationsTemperature2006].

>>>>

The strength of this buffering varies with foliage characteristics mentioned above, and water availability [@davis_microclimatic_2019]. Diurnally, tropical, temperate broadleaf, mixed-conifer and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower maximum temperature and vapor pressure deficit ($VPD$) in the daytime, and warmer higher minimum temperature during nighttime in the understory respectively, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al. 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; Rambo and North, 2009]. Similarly, when compared to neighboring tree crop agricultural plantations, intact tropical forests maintain lower understory peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with LAI and canopy cover [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. On the other hand, open-forests with sparse stands, have lower buffering capacity, thus greater vertical air mixing, greater light permeability through canopy layers and sometimes greater soil heating [@ramboCanopyMicroclimateResponse2009]. Therefore, they maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009]. During the night, pine forests have shown to maintain warmer understory temperatures than above canopy, due to long wave infrared radiation from the soil caused by daytime solar warming [@vonarxSpatiotemporalEffectsForest2012a]. Canopy roughness also affects buffering, turbulent air flow and the canopy boundary layer. As a result, taller trees up to 20m in an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature in the understory. However, beyond a certain tree height, temperature and VPD offset plateaus [@juckerCanopyStructureTopography2018; @hardwickRelationshipLeafArea2015].

@@

While exposed sun leaves are equipped to thermoregulate more efficiently than shaded leaves, they are largely constrained by stomatal conductance; and therefore their ability to photosynthesize as well. During the day, stomtal conductance increases with height and is greater in the upper canopy compared to shaded canopies [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010]. During midday, stomtal closure occurs, when evaporative demand exceeds water transport capacity in exposed leaves [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004]. Thus, stomatal limitation is greater and steeper in exposed canopies than shade leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020]. Similarly, max transpiration rate increases with canopy height [@kosugiSeasonalFluctuationsTemperature2006]. Stomatal apeture, also controls heat dissapation through evaporative cooling [Taiz and Zeiger], where understory leaves tend to open their stomata wider than upper canopy leaves [@kosugiSeasonalFluctuationsTemperature2006].
